---
title: "Chapter 11: Measuring the Spread of Our Data"
draft: false
---

```{r setup}
options("scipen" = 9999, "digits" = 4)
```

This chapter defines the mean absolute deviation (MAD), variance, and
standard deviation of a set of numbers.

## Q1

::: {.callout-note appearance="simple"}

One of the benefits of variance is that squaring the differences makes the
penalties exponential. Give some examples of when this would be a useful
property.

:::

The exponential penalty of the variance can be useful when we expect all of
our data points to be somewhat close together, so that outlying data points
are penalized more severely the further away they are. If error is expensive,
we also want to penalize errors increasingly as they become larger.

## Q2

::: {.callout-note appearance="simple"}

Calculate the mean, variance, and standard deviation for the following
values: `1, 2, 3, 4, 5, 6, 7, 8, 9, 10`.

:::

```{r}
vec <- 1:10
cat("Mean:    ", round(mean(vec), 3), "\n")
cat("MAD:     ", round(mad(vec, center = mean(vec)), 3), "\n")
cat("Variance:", round(var(vec), 3), "\n")
cat("SD:      ", round(sd(vec), 3), "\n")
```

