---
title: "Chapter 12: Monsters and Mixtures"
format: html
---

```{r}
library(rethinking)
```

This chapter is about models where the data likelihood is a "mixture" distribution: the student $t$ distribution is a replacement for Gaussian likelihoods with many outliers; beta-binomial and gamma-Poisson are replacements for binomial and Poisson likelihoods respectively that allow for more heterogeneity in the data; and zero-inflated or hurdle models allow for discrete point masses of probability that arise from different processes. While these models don't give us causal information about what is happening in our data, they can help us get more robust inferences about other things we care about.

We also learned about models for categorical variables with a specific order, the cumulative logit model for outcomes and a monotonicty constraint with dirichlet priors for predictors.

## Exercises

### 12E1

An ordered categorical variable has a natural order for the levels so it makes sense to put them in a specific order, while an unordered (nominal) categorical variable doesn't have a natural ordering. Specifically, we can say that an ordered category variable has levels that increase or decrease in order, even if we can't quantify the amount of change between each level.

An example of an ordered categorical variable is a symptom score, which is designed to be higher if a patient experiences more symptoms or more severe symptoms during an illness, but a change in symptom score from 1 to 2 may not be the same as from 5 to 6. An example of an unordered categorical variable is ABO blood group -- there is no clear order of A, B, AB, and O blood types (although this could arguably be better treated as two indicator variables that interact, one for A antigen and one for B antigen).

### 12E2

An ordered logistic regression uses a *cumulative* logit link function rather than the ordinay logit link. The cumulative logit link uses the logit function applied to the cumulative probability of being in ordered category $k$, i.e., $P(y \leq k)$, instead of the probability that we observe a specific category, i.e., $P(y = k)$.

### 12E3

When count data are zero-inflated, using a model that ignores zero-inflation will make the estimated mean look smaller than it is, because we include many zeros in the estimation of the mean that are not generated by the process with that mean.

### 12E4

A process that might produce over-dispersed counts is disease transmission when particular infected individuals have a chance of being superspreaders. On average, we would expect a particular person to infect a certain number of individuals, but superspreaders infect far more people than most infected individuals, increasing the variance in the case counts we observe.

A process that might show under-dispersion is one where there is a limit to the number of events that can occur per day. For example, if a small clinic has to ration vaccines and can give out at most $k$ per day, if demand is high there will be consistently low variation in the number of vaccines distributed, although it could occasionally be lower so the variance is nonzero.

### 12M1

At this university, the ordinal employee ratings (1-4) were respectively 12, 36, 7, and 41. The cumulative log-odds of each ranking are
$$
\log \frac{P(\text{rating} \leq k)}{1 - P(\text{rating} \leq k)}.
$$

We can compute these like this.

```{r}
counts <- c(12, 36, 7, 41)
probs <- counts / sum(counts)
c_probs <- cumsum(probs)

log(c_probs / (1 - c_probs))
```

The cumulative log-odds of the highest rating is infinite, because everyone was assigned a rating that is at most the highest value.

### 12M2

Let's plot those cumulative ratings.

```{r}
layout(1)
plot(
	NULL, NULL,
	type = "b",
	xlab = "rating", ylab = "cumulative proportion",
	xlim = c(1, length(counts)),
	ylim = c(0, 1),
	xaxt = "n"
)
axis(1, at = 1:4, labels = as.character(1:4))
d <- 0.0075
arrows(
	x = 1:length(counts) - d,
	x0 = 1:length(counts) - d,
	y = c_probs,
	y0 = 0,
	length = 0,
	col = "darkgray",
	lwd = 2
)
arrows(
	x = 1:length(counts) + d,
	x0 = 1:length(counts) + d,
	y = c_probs,
	y0 = c_probs - probs,
	length = 0,
	lwd = 2, col = rethinking::rangi2
)
points(
	1:length(counts),
	c_probs,
	type = "b", pch = 16, cex = 1.1
)
```

The points and the gray bars show the cumulative probability of a rating of $k$ or higher, while the blue bars show the difference in probability between rating $k$ and rating $k-1$.

### 12M3

Earlier in the chapter, the zero-inflated Poisson likelihood was given as
$$
(1-p)\left(\frac{\lambda^y\exp(-y)}{y!}\right).
$$

We can write a similar zero-inflated binomial likelihood, we just can't use the letter $p$ twice. So let $\pi$ be the probability that an observation is a zero from the process that only produces zeroes, and let $n$ and $p$ be the regular binomial parameters. Then,

$$
P(y \mid \pi, n, p) = \left(\pi \times 0 \right) + \left((1-\pi) \times \left( n \atop y \right) p ^ y\left(1 - p\right)^{n-y}\right),
$$
if $y=1$, because all ones have to come from the binomial part, and
$$
P(y \mid \pi, n, p) = \pi + \bigg((1- \pi) \times (1-p)^n\bigg)
$$
if $y=0$ because a zero could be either from the zero-generating process or from the binomial part.
