---
title: "SR Chapter 9: Markov Chain Monte Carlo"
---

```{r setup, include = TRUE}
#| message: false
#| warning: false
library(rethinking)
library(dagitty)
library(cmdstanr)
```

This chapter covers Markov Chain Monte Carlo, Gibbs sampling, Hamiltonian
Monte Carlo, and the HMC implementation in `rethinking`.


## Exercises

### 9E1

The simple Metropolis algorithm requires that the proposal distribution must
be symmetric.

### 9E2

Gibbs sampling achieves better efficiency than the simple Metropolis algorithm
by using conjugate prior distributions to adapt the proposal distribution
as the parameters change. This allows a Gibbs sampler to make intelligent
jumps around the posterior distribution, rather than blindly guessing. However,
Gibbs samplers require the use of these conjugate priors, and often fail
to retain their efficiency for high-dimensional, multilevel parameter spaces.

### 9E3

Hamiltonian Monte Carlo cannot handle discrete parameters, because the gradient
is not continuous and the physics simulation that generates proposals will
not be able to move along a discrete parameter space.

### 9E4

The effective number of samples is the number of samples from the posterior
adjusted for the autocorrelation between successive samples. Autocorrelated
samples do not provide as much information as independent samples, and so
multiple highly autocorrelated draws give us less information than the reported
sample size would have if the draws were independent. Conversely, anticorrelated
samples provide more information than the corresponding number of independent
samples.

### 9E5

If the chains are mixing correctly, $\hat{R}$ should approach 1.

### 9E6

A good trace plot should show constant movement of each chain in the model, and
the chains should cross each other constantly. This indicates that all chains
are exploring the posterior distribution without getting stuck, and they are
mixing well. A bad trace plot would have chains which are separated from each
other, indicating that they are all exploring different areas of the posterior
distribution without mixing; or some chains could get stuck near particular
parameter values, indicating that chain is failing to correctly explore.

### 9E7

A good trace rank plot will look similar to a good trace plot, but we will
be able to see whether the chains are mixing more easily because each chain
will be constantly change ranks at the next sampling iteration.

### 9M1

First we need to restimate the terrain ruggedness model using a uniform prior
distribution for the `sigma` parameter.

```{r 9m1.1}
data("rugged")
d <- rugged
d$log_gdp <- log(d$rgdppc_2000)
dd <- subset(
	d,
	subset = complete.cases(rgdppc_2000),
	select = c(log_gdp, rugged, cont_africa)
)
dd$log_gdp_std <- dd$log_gdp / mean(dd$log_gdp)
dd$rugged_std <- dd$rugged / max(dd$rugged)
dd$cid <- ifelse(dd$cont_africa == 1, 1, 2)

m_9m1 <- rethinking::ulam(
	flist = alist(
		log_gdp_std ~ dnorm(mu, sigma),
		mu <- a[cid] + b[cid] * (rugged_std - 0.215),
		a[cid] ~ dnorm(1, 0.1),
		b[cid] ~ dnorm(0, 0.3),
		sigma ~ dunif(0, 1)
	),
	data = dd,
	chains = 1
)

rethinking::precis(m_9m1, depth = 2)
```

The mean and credible intervals for all the parameters are very similar,
although the ESS is lower, indicating that sampling was less efficient using a
uniform prior. However, since we collected enough samples and we have enough
data, the prior does not have a strong influence on estimating our parameters.
Notably, the estimated value of sigma in the previous model was within the
constraints of the new prior we used, so even though a uniform prior imposes
constraints on the possible values of sigma, we didn't actually run into those
boundaries in practice. If our sigma value were larger than one, we would
expect this new posterior to look different.

### 9M2

Now we'll put the prior for `sigma` back to exponential but we'll change the
prior the slopes.

```{r 9m2.1}
m_9m2 <- rethinking::ulam(
	flist = alist(
		log_gdp_std ~ dnorm(mu, sigma),
		mu <- a[cid] + b[cid] * (rugged_std - 0.215),
		a[cid] ~ dnorm(1, 0.1),
		b[cid] ~ dexp(0.3),
		sigma ~ dexp(1)
	),
	data = dd,
	chains = 1
)

rethinking::precis(m_9m2, depth = 2)
```

Now, of course, the estimated slopes have changed a bit. Because of the
exponential prior we specified, the slopes that we estimate MUST be positive.
In the previous model, `b[2]` was negative and now it can't be. Notably, our
ESS has decreased further from the original model as well.

### 9M3

We'll rerun the terrain ruggedness model with the original prior for multiple
different numbers of warm-up samples. Although this will take a bit to run
because of the compilation step, I'll go do something else while it's running
so it's not too annoying.

```{r 9m3.1}
fit_9m3 <- function(warmup) {
	model <- rethinking::ulam(
		flist = alist(
			log_gdp_std ~ dnorm(mu, sigma),
			mu <- a[cid] + b[cid] * (rugged_std - 0.215),
			a[cid] ~ dnorm(1, 0.1),
			b[cid] ~ dnorm(0, 0.3),
			sigma ~ dexp(1)
		),
		data = dd,
		chains = 1,
		warmup = warmup,
		iter = 500 + warmup
	)
	
	return(model)
}

get_ess <- function(ulam_fit) {
	prc <- rethinking::precis(ulam_fit, depth = Inf)
	res <- prc$ess_bulk
	names(res) <- rownames(prc)
	return(res)
}

wu_iters <- c(25, 50, 100, 250, 500)
res <- purrr::map(
	.x = wu_iters,
	.f = \(w) w |>
		fit_9m3() |>
		get_ess()
)

res_df <- do.call(rbind, res)
plot_colors <- RColorBrewer::brewer.pal(5, "Dark2")
plot(
	NULL, NULL,
	xlim = c(25, 500),
	ylim = c(0, 1000),
	xlab = "Number of warmup samples",
	ylab = "Bulk ESS"
)
for (i in 1:nrow(res_df)) {
	lines(
		wu_iters,
		res_df[, i],
		col = plot_colors[[i]],
		lwd = 1.25
	)
	points(
		wu_iters,
		res_df[, i],
		col = plot_colors[[i]],
		pch = i,
		cex = 1.1
	)
}
legend(
	x = "bottomright",
	legend = colnames(res_df),
	col = plot_colors,
	pch = 1:5,
	cex = 1.05,
	lwd = 1.25,
	ncol = 3,
	title = "Parameter"
)
	

```

Based on this quick experiment, it appears that around 250 warmup samples is
enough to get consistent ESS for all parameters in the model. However, for a
more complicated model we would likely need more warmup iterations.

### 9H1

First we'll run the model specified by the book.

```{r 9H1.1}
mp <- rethinking::ulam(
	alist(
		a ~ dnorm(0, 1),
		b ~ dcauchy(0, 1)
	),
	data = list(y = 1),
	chains = 1
)
```

Now we'll look at the precis.

```{r 9H1.2}
rethinking::precis(mp)
```

And the trace plots.

```{r 9H1.3}
rethinking::traceplot(mp)
```

We can see that the traceplot and summary for the parameter `a` are normal,
but the parameter `b` has an inflated variance, lower ESS, and kind of crazy
looking traceplot. This is because the normal distribution is typically very
well-behaved, but the Cauchy distribution is often problematic. The Cauchy
distribution has no expected value and very thick tails, so we can see that
when the HMC sampler is exploring the density of parameter `b`, it sometimes
goes into regions of the distribution which are quite far from zero, because
they are, in general, much likelier than the corresponding regions for the
parameter `a`. That means that we would need more samples and more data
to accurately sample a Cauchy parameter than a Normal parameter.

These properties of the Cauchy distribution are partially why Stan developer
Michael Betancourt recommends not using Cauchy priors at all, despite their
popularity for a period of time.

### 9H2

For this problem, we'll refit models `m5.1`, `m5.2`, and `m5.3` from the chapter 5 divorce rate example. First we need to set up the data. Note that we can't use McElreath's code *exactly* anymore because Stan now complains about dots in the variable names of `d` if we include all the variables from `WaffleDivorce` in the `ulam()` data list.

```{r waffledivorce setup}
data("WaffleDivorce")
d <- list()
d$D <- rethinking::standardize(WaffleDivorce$Divorce)
d$M <- rethinking::standardize(WaffleDivorce$Marriage)
d$A <- rethinking::standardize(WaffleDivorce$MedianAgeMarriage)
```

Now we'll refit the three models using `ulam()`.

```{r m5.1 to m5.3}
m5.1 <- rethinking::ulam(
	alist(
		D ~ dnorm(mu, sigma),
		mu <- a + bA * A,
		a ~ dnorm(0, 0.2),
		bA ~ dnorm(0, 0.5),
		sigma ~ dexp(1)
	),
	data = d,
	log_lik = TRUE
)

m5.2 <- rethinking::ulam(
	alist(
		D ~ dnorm(mu, sigma),
		mu <- a + bM * M,
		a ~ dnorm(0, 0.2),
		bM ~ dnorm(0, 0.5),
		sigma ~ dexp(1)
	),
	data = d,
	log_lik = TRUE
)

m5.3 <- rethinking::ulam(
	alist(
		D ~ dnorm(mu, sigma),
		mu <- a + bM * M + bA * A,
		a ~ dnorm(0, 0.2),
		bM ~ dnorm(0, 0.5),
		bA ~ dnorm(0, 0.5),
		sigma ~ dexp(1)
	),
	data = d,
	log_lik = TRUE
)
```

Now we'll compare the models using PSIS.

```{r m5.x compare}
rethinking::compare(
	m5.1, m5.2, m5.3,
	sort = "PSIS",
	func = PSIS
)
```

Notably, McElreath's PSIS output is formatted so that *lower PSIS is better*, which is the opposite of the ELPD metric used by the posterior and loo developers. Keeping this in mind, we see that the two models containing age at marriage both beat the model with only the marriage rate. These two models perform similarly, with a low standard error for the difference in their PSIS criterion. Notably, the difference between the best model and the worst model also has a very low ratio of the difference to the standard error, so even these two models are not *very* different.

### 9H3

First, we need to use the code from Chapter 6 to simulate some leg length data.

```{r 9h3 setup}
N_leg <- 100L
set.seed(909)
d_leg <- tibble::tibble(
	height = rnorm(N_leg, 10, 2),
	leg_prop = runif(N_leg, 0.4, 0.5),
	leg_left = leg_prop * height + rnorm(N_leg, 0, 0.02),
	leg_right = leg_prop * height + rnorm(N_leg, 0, 0.02)
)

head(d_leg)
```

Now we'll refit the first model and fit the new model with the constrained prior, this time using `ulam()` for both fits.

```{r 9h3 models}
m5.8s <- rethinking::ulam(
	alist(
		height ~ dnorm(mu, sigma),
		mu <- a + bl * leg_left + br * leg_right,
		a ~ dnorm(10, 100),
		bl ~ dnorm(2, 10),
		br ~ dnorm(2, 10),
		sigma ~ dexp(1)
	),
	data = d_leg,
	chains = 4L,
	cores = 4L,
	start = list(a = 10, bl = 0, br = 0.1, sigma = 1),
	log_lik = TRUE
)

m5.8s2 <- rethinking::ulam(
	alist(
		height ~ dnorm(mu, sigma),
		mu <- a + bl * leg_left + br * leg_right,
		a ~ dnorm(10, 100),
		bl ~ dnorm(2, 10),
		br ~ dnorm(2, 10),
		sigma ~ dexp(1)
	),
	data = d_leg,
	chains = 4L,
	cores = 4L,
	constraints = list(br = "lower=0"),
	start = list(a = 10, bl = 0, br = 0.1, sigma = 1),
	log_lik = TRUE
)
```

OK, we get some diagnostic warnings from Stan about both of these models, but for now we'll ignore those since the book doesn't mention anything about it. (It makes sense because these models are kind of poorly parametrized anyway, which is the point of these models.) But now we need to compare the posterior distributions for the parameters in both models, so we'll just plot both.

```{r 9h3 compare}
plot(rethinking::coeftab(m5.8s, m5.8s2))
```

The models are not all that different, the `br` parameter was already positive so adding the lower bound did not completely flip the sign of this parameter. However, since we assumed a higher prior likelihood of a positive coefficient, the `br` value was higher for the second model. Since the `br` and `bl` parameters are so strongly correlated, this forces the `bl` parameter to be smaller than it was before, to basically make up for the difference. The two parameters are nearly perfectly correlated so if we force one to shift upwards, the other has to shift downwards.

### 9H4

Now we need to compare the two models with PSIS.

```{r 9h4 compare}
rethinking::compare(
	m5.8s, m5.8s2,
	sort = "PSIS", func = PSIS
)
```

The original model has more effective parameters, slightly. But the difference is not really notable. However, the model that we imposed a constrained prior on has a lower effective parameter size, because we've excluded an entire region from being traversed by the sampler for a specific parameter, so we are more confident in that parameter.

### 9H5

For this problem, we'll modify the Metropolis algorithm code from the chapter to handle the case that the island populations have a different distribution than the island numbers.

```{r 9h5 code}
# First generate the 10 island populations
set.seed(375)
populations <- floor(runif(10, 1, 100))

# Now set up the num_weeks, positions and current variables, nothing
# changes in this part
num_weeks <- 1e5
positions <- rep(0, num_weeks)
current <- 10

# Most of the loop stays the same, we generate the proposal the same way
for (i in 1:num_weeks) {
	## Record current position
	positions[i] <- current
	
	## Generate proposal
	proposal <- current + sample(c(-1, 1), size = 1)
	
	## Make sure he loops around the archipelago
	if (proposal < 1) proposal <- 10
	if (proposal > 10) proposal <- 1
	
	## Next calculate the probability of moving based on the relative
	## populations, which are not equal to the positions
	prob_move <- populations[proposal] / populations[current]
	current <- ifelse(runif(1) < prob_move, proposal, current)
}
```

We only had to make a small adjustment to the code. Let's check to make sure we're spending time proportional to the island populations.

```{r 9h5 plots}
layout(matrix(c(1, 2), nrow = 1))

k <- 1000
plot(
	1:k, positions[1:k], type = "l",
	xlab = "Week",
	ylab = "Island"
)

plot(
	prop.table(table(positions)),
	xlab = "Island",
	ylab = "Relative frequency of time on island"
)

points(
	x = 1:10,
	y = populations / sum(populations),
	col = "red",
	pch = "-",
	cex = 3
)

legend(
	"topright",
	col = "red", pch = "-",
	legend = "Island population / total population"
)

layout(1)
```

The left plot shows that we're bouncing randomly between the islands, sometimes going back and forth between the same two, occasionally moving many times in a row, and occasionally staying on the same island. The right plot shows that we're getting the right approximation -- the amount of time we spend on each island is roughly the same as that island's proportion of the total population. For reference I'll print all of the island populations below.

```{r 9h5 pops}
populations
```

### 9H6

For this exercise we'll modify the Metropolis algorithm code from the chapter to write our own MCMC example for the globe tossing data and model. Following McElreath's sketch from the beginning of section 9.2, we have that

- The "islands" are the possible values of the parameter $p$, the proportion of the globe that is water;
- The "population sizes" are the probabilities of each value of $p$ after we observed the globe tossing values; and
- The "weeks" are not interpretable as time values but are the samples we take from the posterior distribution of $p$.

From chapter 2, we have 6 water samples and 3 land samples. We'll set our initial value therefore as the MLE for the proportion of water, $6 / 9 \approx 0.67$.

```{r 9h6 code}
# Initialization
set.seed(376)
num_samples <- 1e5
p_samples <- rep(0, num_weeks)
current <- 0.67

# Most of the loop stays the same, we generate the proposal the same way
for (i in 1:num_samples) {
	## Record current position
	p_samples[i] <- current
	
	## Generate proposal -- there are multiple ways to do this, but we'll
	## just generate a new value that's within a certain distance of p. Then
	## if p is outside the bounds [0, 1], we'll set it to the boundary.
	proposal <- current + runif(1, -0.25, 0.25)
	if (proposal < 0) proposal <- 0
	if (proposal > 1) proposal <- 1
	
	## Next calculate the likelihood of the current and proposed values
	proposal_p <- dbinom(6, 9, proposal)
	current_p <- dbinom(6, 9, current)
	
	## Move?
	prob_move <- proposal_p / current_p
	current <- ifelse(runif(1) < prob_move, proposal, current)
}
```

Now we can look at the estimated density of the posterior distribution of $p$ that we've sampled from.

```{r 9h6 plots}
p_dens <- density(p_samples)
p_mode <- p_dens$x[which.max(p_dens$y)]
plot(
	p_dens, main = "",
	lwd = 2
)
abline(v = p_mode, lty = 2, col = "red", lwd = 1.5)
text(paste0("MAP = ", round(p_mode, 3)), x = p_mode - 0.05, y = 1.0)
```

The maximum *a posteriori* estimate from our sampled values is `r round(p_mode, 3)`, which is quite close to the true estimated value somewhere near $0.73$. Having more data samples would improve this a lot; adding more MCMC samples probably wouldn't. But let's look at the trace plot just to make sure.

```{r 9h6 plot results}
k <- num_samples / 10
plot(x = 1:k, y = p_samples[1:k], type = "l", xlab = "draw", ylab = "p")
```

We can see that the chain moves up and down randomly (for these draws, which are the first `r k`). So it seems that our estimator didn't get stuck anywhere and we can probably trust it.

### 9H7

Now we want to write our own HMC algorithm for the globe tossing data. We need to write functions for the likelihood and gradient. We know that the likelihood is

$$L(W \mid p, n) = \left({n}\atop{W}\right) p^W (1-p)^{n-W},$$
and thus the log likelihood is
$$\ell(W \mid p, n) = \log\left({n}\atop{W}\right) + W\log p + (n-w)\log(1-p).$$

Now we need to get the derivative of the log-likelihood, which is
$$
\frac{d\ell}{dp} = \frac{w-np}{p(1-p)}.
$$

Let's implement both of these in R, ignoring the fact that the log-likelihood can already be obtained as `dbinom(..., log = TRUE)`.

```{r 9h7 functions}
U <- function(p, W = 6, n = 9) {
	U <- log(choose(n, W)) + W * log(p) + (n - W) * log(1 - p)
	return(-U)
}

U_gradient <- function(p, W = 6, n = 9) {
	G <- (W - n * p) / (p * (1 - p))
	return(-G)
}
```

I checked these against `dbinom(...)` and a simple central difference differentiation scheme to make sure they were giving the same answers as the built-in function. Note that the difference between the derivatives gets larger as $h$ increases as we expect, but for suitably small $h$ there is no difference between our method and the numerical derivative for the function that we know is correct.

```{r 9h6 functions test}
test_p <- c(0.05, 0.25, 0.5, 0.75, 0.95)

our_U <- U(test_p)
r_U <- dbinom(6, 9, test_p, TRUE) * -1
all.equal(our_U, r_U)

h <- 0.00001
r_dU <- (dbinom(6, 9, test_p + h, TRUE) - dbinom(6, 9, test_p - h, TRUE)) / (2*h)
our_dU <- U_gradient(test_p) * -1
all.equal(our_dU, r_dU)
```

Now that we have the functio and the gradient programmed, we can implement the actual HMC routune.

```{r 9h7 run HMC}
set.seed(377)
Q <- list()
Q$q <- c(0.67)
pr <- 0.3
step <- 0.03
L <- 11
n_samples <- 1e3
Q_samples <- numeric(n_samples)
time <- 1:n_samples
path_col <- rethinking::col.alpha("black", 0.5)
for (i in 1:n_samples) {
	# Do the samples and save the output
	Q <- rethinking::HMC2(U, U_gradient, step, L, Q$q)
	Q_samples[i] <- Q$q
}
```

I wanted to figure out how to make a plot like Figure 9.5 on page 272, but the code and plotting stuff wasn't explained enough for me to do it in a reasonable time, so we'll just look at the trace plot and density like usual. Note that I tried running more samples, but I kept getting weird internal errors for `HMC2` so I don't think it's terribly robust, and the number of samples that you can succesfully complete probably is dependent on the starting seed.

```{r 9h7 plots}
layout(matrix(c(1, 2), nrow = 1))
plot(
	time, Q_samples, type = "l",
	xlab = "Draw", ylab = "p"
)

dens_hmc <- density(Q_samples)
hmc_mode <- dens_hmc$x[which.max(dens_hmc$y)]
plot(
	dens_hmc, main = "",
	lwd = 2
)
abline(v = hmc_mode, lty = 2, col = "red", lwd = 1.5)
text(paste0("MAP = ", round(p_mode, 3)), x = p_mode - 0.075, y = 1.0)
```

For this example, we got that the MAP after `r n_samples` samples was `r round(hmc_mode, 3)` which is similar to the truth and similar to our Metropolis MCMC result.

<!-- END OF FILE -->
